---
title: "Peer Assessment 1 —  Activity monitoring data"
author: "LLCP"
date: "June 14, 2015"
output: html_document
---

## Loading and preprocessing the data

The data has been preprocess using dplyr library, dplyr provides a flexible grammar of data manipulation

```{r, warning=FALSE, message=FALSE}
library(dplyr)
data <- read.csv(file = "activity.csv", as.is = TRUE)
data <- tbl_df(data)
```


## What is mean total number of steps taken per day?

First, I calculate the total number of steps group by day

```{r, warning=FALSE, message=FALSE}
data_question_1 <- data %>% 
                         group_by(date) %>%
                         summarise(stepsByDay= sum(steps, na.rm = T))

```

Then, I plot the histogram for the **data_question_1** already compute: 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
qplot(stepsByDay, data=data_question_1, geom= "histogram")
```

Finally, I calculate and report the mean and median of the total number of steps taken per day: 

```{r, warning=FALSE, message=FALSE}
mean(data_question_1$stepsByDay)
median(data_question_1$stepsByDay)
```

## What is the average daily activity pattern?

First, I calculate the mean number of steps group by interval

```{r, warning=FALSE, message=FALSE}
data_question_2 <- data %>% 
                         group_by(interval) %>%
                         summarise(stepsByInterval= mean(steps, na.rm = T))

```

Then, I plot the 5-minute interval average number of steps taken, to show the daily pattern:

```{r, warning=FALSE, message=FALSE}
qplot(interval, stepsByInterval, data=data_question_2, geom=c("point", "smooth"))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 

```{r, warning=FALSE, message=FALSE}
data_question_2 %>% filter( stepsByInterval == max(stepsByInterval))

```

## Imputing missing values

The presence of missing days may introduce bias into some calculations or summaries of the data. I will calculate and report the total number of missing values in the dataset

```{r, warning=FALSE, message=FALSE}
sum(is.na(data$steps))
```

The strategy for filling the missing values in the dataset is filling that value with the mean for that 5-minute interval. Let's do it! I already have the mean 5-minute interval in the **data_question_2** dataset     

```{r, warning=FALSE, message=TRUE}
data_question_3 <- left_join(data, data_question_2) %>%
                  mutate(fill.steps = ifelse(is.na(steps), stepsByInterval, steps)) %>%
                  select(date, interval, fill.steps)
```

I will make a histogram of the total number of steps taken each day 

```{r, warning=FALSE, message=FALSE}
qplot(fill.steps, data=data_question_3, geom= "histogram")
```

Now, I calculate and report the mean and median total number of steps taken per day.

```{r, warning=FALSE, message=FALSE}
mean(data_question_3$fill.steps)
median(data_question_3$fill.steps)
```

Do these values differ from the estimates from the first part of the assignment? Sure, lets look how much they differ

```{r, warning=FALSE, message=FALSE}
mean(data_question_1$stepsByDay) - mean(data_question_3$fill.steps)
median(data_question_1$stepsByDay) - median(data_question_3$fill.steps)
```

**What is the impact of imputing missing data on the estimates of the total daily number of steps? 
The impact is huge! Now, this guys looks like a very sedentary man.**

## Are there differences in activity patterns between weekdays and weekends?

I create a new variable in the dataset with two values – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, warning=FALSE, message=FALSE}
data_question_4 <- data %>% mutate(day =weekdays(as.Date(data$date))) %>%
                  mutate(weekday = ifelse((day == "Sunday" | day == "Saturday"),"weekend", "weekday")) 

data_question_4 <- data_question_4 %>% 
                  select(steps, date, interval, weekday) %>%
                  group_by(weekday, interval) %>%
                  summarise(stepsByInterval= mean(steps, na.rm = T))


```

Let's see if there is any difference in the between weekday and weekend

```{r, warning=FALSE, message=FALSE}
qplot(interval, stepsByInterval,
      data=data_question_4, facets=.~weekday,
      geom=c("point","smooth"), color=weekday)
```
